---
title: "Analysis of Professor salaries"
author: "Nick Cianci"
date: "November 27, 2017"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, warning=FALSE, message=FALSE}
#Load in relevant packages
library('DataComputing')
library('knitr')
library('mosaic')
library('readr')
library('dplyr')
```

#The Data
The dataset I chose was a csv file of Salaries of Professors with PhD's. I obtained the dataset from the webstite https://vincentarelbundock.github.io/Rdatasets/datasets.html 
```{r, warning=FALSE, message=FALSE}
#Read in Professor's salary data and look at the structure of the data

raw_data2 <- read_csv('https://raw.githubusercontent.com/vincentarelbundock/Rdatasets/master/csv/car/Salaries.csv')

str(raw_data2)

```

By looking at the structure of the data, you can see that their are 397 observations with 7 attributes however the first attribute is simply an index of the records and was not used throughout the project. The attributes used were rank, discipline, yrs.since.phd, yrs.service, sex, and salary. Rank, discipline and sex are characters while yrs.since.phd, yrs.service, and salary are integers.



Lets take a look at the first 6 records in the dataset
```{r}
#A quick look at the data
kable(head(raw_data2))
```

#Creating a new dataframe
I didn't have any use for the first attribute (X1), so I created a new dataframe that is exactly the same data but with the first row dropped.
```{r}
#drop the first row
rank <- raw_data2$rank
discipline <- raw_data2$discipline
yr_since_phd <- raw_data2$yrs.since.phd
yrs_service <- raw_data2$yrs.service
sex <- raw_data2$sex
salary <- raw_data2$salary



#create a new dataset without the first column
df2 <- data.frame(rank, discipline, yr_since_phd, yrs_service,sex, salary)
kable(head(df2))
```

#Does longevity mean a bigger payday?
I was interested in exploring whether or not the amount of years since obtaining a PhD or the amount of years of service affected the salary size for Professors. I created multiple graphs in an attempt to uncover any hidden trends. 

```{r}
#scatter plot of the relationships between salary and years since obtaining a phd, colored by different rank and faceted by discipline
ggplot(data= df2 ,aes(x=yr_since_phd, y=salary))+geom_point()+aes(colour = rank)+facet_wrap(~discipline) 

#line plot of the relationships between salary and years since obtaining a phd, colored by different rank and faceted by discipline
ggplot(data= df2 ,aes(x=yr_since_phd, y=salary))+geom_line()+aes(colour = rank)+facet_wrap(~discipline) 
```



```{r}
#scatterplot of the relationships between salary and years of service, colored by different rank and faceted by discipline
ggplot(data= df2 ,aes(x= yrs_service , y=salary))+geom_point()+aes(colour = rank)+facet_wrap(~discipline) 

#line plot of the relationships between salary and years of service, colored by different rank and faceted by discipline
ggplot(data= df2 ,aes(x=yrs_service, y=salary))+geom_line()+aes(colour = rank)+facet_wrap(~discipline) 


```


What I can conclude from these graphs is that generally, the more years of service correlates to a higher salary. The graphs for years since obtaining a PhD were strikingly similar to the graphs for years of service and thus have the same conclusions. I also concluded that the rank of the professor (Associate, Assistant, Professor) is a much better indicator of salary than the years of service or years since PhD, as one would expect.Their doesn't seem to be any significant difference in salary distribution based on the discipline of the professors. 

#Do Male Professors make more than Female Professors?
Something else, I wanted to explore with this dataset was if Professor's sex had any bearing on the salary that they receive. I did this by splitting the dataset into male and female professors. I simply got the maximum, minimum, and the average of the salaries for both males and females then compared them via a bar plot. 
```{r}
#Filters the dataframe for only males
male_sal <- df2 %>%
  filter(sex == 'Male')

#Max salary of males
max(male_sal$salary)

#Min salary for males
min(male_sal$salary)

#Average salary for males
mean(male_sal$salary)


#Filters the dataframe for only females
female_sal <- df2 %>%
  filter(sex == 'Female')

#Max salary for females
max(female_sal$salary)

#Min salary for females
min(female_sal$salary)

#Average Salary for females
mean(female_sal$salary)


```


```{r}

#bar plot of the relationships between salary and years since obtaining a phd, colored by different rank and faceted by discipline
ggplot(data= df2 ,aes(x=rank, y= salary))+geom_point()+aes(colour = discipline)+facet_wrap(~sex) 
```

As the calculations and the plot suggest, Male professors seem to make higher salaries than Female professors. The Male professor makes $14,088 more on average than their female counterparts. Meanwhile,the highest recorded salary also belonged to a male professor coming in at $231,545, an eyepopping $70,444 more than the largest female professor salary. 
